# Practice MR Eggar - 24OCT25

Experiment:

-   Exposure: Menarche timing - from a supplementary file.

-   Outcome: T2D GWAS - from Felix.

#### Univariate MR Eggar.

Notes:

-   The univariable MR-Egger method is the same as the IVW method using weighted linear regression except the intercept term is estimated rather than being set to zero.

-   Testing whether the intercept term is equal to zero is equivalent to testing for directional pleiotropy and the validity of the InSIDE assumption.

-   The genetic associations with the risk factor bX1 and outcome bY must be orientated with respect to the risk increasing or decreasing allele of the risk factor.

Equations for Univariate MR Eggar.

Under the MR-Egger model, multiplicative random-effects should be used as the presence of pleiotropy will lead to overdispersion.

Since the residual standard error is estimated, we use the t-distribution with J − 2 degrees of freedom for inference.

``` r
#Orientation of the genetic associations
bY<-ifelse(bX1>0, bY, bY-1)
bX1<-abs(bX1)

#Causal estimate
thetaUE = summary(lm(bY~bX1, weights=bYse^-2))$coef[2]
se_thetaUE.random = summary(lm(bY~bX1, weights=bYse^-2))$coef[2,2]/ min(summary(lm(bY~bX1, weights=bYse^-2))$sigma,1)
lb_thetaUE = thetaUE - qt(0.975,df=length(bX1)-2)*se_thetaUE.random 
ub_thetaUE = thetaUE + qt(0.975,df=length(bX1)-2)*se_thetaUE.random 
p_thetaUE = 2*(1-pt(abs(thetaUE/se_thetaUE.random),df=length(bX1)-2)) 

#Test for directional pleiotropy 
interUE = summary(lm(bY~bX1, weights=bYse^-2))$coef[1] 
se_interUE.random = summary(lm(bY~bX1, weights=bYse^-2))$coef[1,2]/ min(summary(lm(bY~bX1, weights=bYse^-2))$sigma,1) 
p_interUE = 2(1-pt(abs(interUE/se_interUE.random),df=length(bX1)-2))
```

Definitions of the variables

-   bY: The associations of the genetic variants with the outcome.

-   bYse: Standard error of the associations of the genetic variants with the outcome.

-   bX1: The associations of the genetic variants with the risk factors are denoted bXk with standard error bXk se, where k = 1, ..., K.

**Reading merged and harmonized file**

```{r}
library(tidyverse, quietly = TRUE)

merged <- read.delim("data/menarche_t2d_merged.tsv", sep = "\t",
                      stringsAsFactors = FALSE)
```

Univariate MR

```{r}
#Assigning variables
bY <- merged$beta_outcome_harmonised
bX1 <- merged$beta_exposure
bYse <- merged$se_outcome

#Orientation of the genetic associations
bY<-ifelse(bX1>0, bY, bY-1)
bX1<-abs(bX1)


```

```{r}
#Causal estimate
thetaUE = summary(lm(bY~bX1, weights=bYse^-2))$coef[2]
se_thetaUE.random = summary(lm(bY~bX1, weights=bYse^-2))$coef[2,2]/ min(summary(lm(bY~bX1, weights=bYse^-2))$sigma,1)
lb_thetaUE = thetaUE - qt(0.975,df=length(bX1)-2)*se_thetaUE.random 
ub_thetaUE = thetaUE + qt(0.975,df=length(bX1)-2)*se_thetaUE.random 
p_thetaUE = 2*(1-pt(abs(thetaUE/se_thetaUE.random),df=length(bX1)-2)) 

#Test for directional pleiotropy 
interUE = summary(lm(bY~bX1, weights=bYse^-2))$coef[1] 
se_interUE.random = summary(lm(bY~bX1, weights=bYse^-2))$coef[1,2]/ min(summary(lm(bY~bX1, weights=bYse^-2))$sigma,1) 
p_interUE = 2*(1-pt(abs(interUE/se_interUE.random),df=length(bX1)-2))
```

Scatter plot

```{r}
#IVW MR
# Weighted linear regression IVW MR
ivw_model <- lm(beta_outcome_harmonised ~ beta_exposure - 1,
                data = merged,
                weights = 1 / se_outcome^2)

# IVW causal estimate
thetaIVW <- summary(ivw_model)$coef[1]

```

```{r}
ggplot(merged, aes(x = beta_exposure, y = beta_outcome_harmonised)) +
  geom_point(size = 2, color = "blue") +  
 # geom_abline(slope = thetaUE, color = "pink", linetype = "dashed", size = 1) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, # MR eggar univariate
              aes(weight = bYse^-2),
              color = "red", linetype = "dashed", linewidth = 1) +
  #geom_abline(slope = thetaIVW, intercept = 0, color = "green", linetype = "dashed", size = 1) +  # IVW MR line 
  geom_smooth(method = "lm", formula = y ~ x - 1, se = TRUE, aes(weight = bYse^-2),
              color = "yellow", linetype = "dashed", linewidth = 1)
  theme_minimal() +
  labs(
    title = "Mendelian Randomization Scatter Plot",
    x = "SNP effect on exposure (bX1)",
    y = "SNP effect on outcome (bY)"
  )
```

#### 
